---
title:  
    "Day6 exercise solutions"
date: 
    "Oct. 21st, 2024"
author:  
    "Ali Movasati"  
output:  
    pdf_document:
        latex_engine: xelatex
---


```{r global options}

# Set global code chunk options
knitr::opts_chunk$set(warning = FALSE)

```


```{r libraries}


# load required libraries
library(skimr)
library(ggplot2)
library(ggpubr)
library(magrittr)
library(tidyr)
library(dplyr)
library(tibble)
library(lme4)
library(lattice)


# define functions
`%notin%` <- Negate(`%in%`)


```

# Problem 1

```{r}

# read in the data

hearing <- read.table(file = "/Users/alimos313/Documents/studies/phd/university/courses/stat-modelling/StatModelEx/day6/data/hearing.txt", header = T, sep = "\t")

```

## 1.A)

```{r}

# print descriptive 
skim(hearing)

table(hearing$ListID)

# prepare data

hearing %<>% mutate(ListID = as.factor(ListID), SubjectID = as.factor(SubjectID))

# plot a box-plot for visualization

hearing %>% ggplot(aes(x = ListID, y = Hearing, color = ListID)) +
        geom_boxplot() +
        geom_jitter(width = 0.1, aes(color = factor(ListID)), size = 2) +
        theme_minimal()


```

## 1.B)

```{r}

lm_simple <- lm(Hearing ~ 1 + ListID, data = hearing)

sum_model_simple <- summary(lm_simple)

print(sum_model_simple)

```

### << comments >>

Only `r round(sum_model_simple$adj.r.squared*100, digits = 2)`% of variability in hearing measures are explained by different lists

We have enough evidence to state that the mean hearing score for List 3 and 4 are different than list 1, while for list 2 we cannot state that!

## 1.C)

```{r}

# fit the mixed model
lm_mixed <- lmer(Hearing ~ 1 + ListID + (1|SubjectID), data = hearing)

sum_model_mixed <- summary(lm_mixed)

print(sum_model_mixed)

# fit the model without ListID (null model)
lm_mixed_null <- lmer(Hearing ~ (1 | SubjectID), data = hearing)

sum_model_mixed_null <- summary(lm_mixed_null)

# Likelihood ratio test
anova(lm_mixed_null, lm_mixed)


```

## 1.D)

### << comments >>

Both models indicate that there are significant differences between the mean hearing score of word lists, therefore
background noise changes the difficulty level of these test!

# Problem 2


```{r}

# read in the data

termites <- read.table(file = "/Users/alimos313/Documents/studies/phd/university/courses/stat-modelling/StatModelEx/day6/data/termites.txt", header = T, sep = " ")

```


## 2.A)

```{r}

# explore the data
skim(termites)


```

## 2.B)

```{r}

# get the data into long format using tidyr::pivot_longer

termites %<>% pivot_longer(cols = starts_with("day"), names_to = "day", values_to = "n.termites") %>%
        mutate(day = sapply(str_split(day, pattern = "day"), "[", 2))


head(termites)



# xyplot to show termite survival over time by dose

xyplot(n.termites ~ day | factor(dose), 
       data = termites_long, 
       groups = factor(dish),
       type = c("o"), # Points and regression lines
       auto.key = TRUE,    # Automatically create a legend
       main = "Survival of Termites Over Days by Dose",
       xlab = "Day",
       ylab = "Number of Surviving Termites",
       layout = c(1, 2))

# Boxplot of n.termites grouped by dose
bwplot(factor(dose) ~ n.termites, 
       data = termites_long,
       main = "Boxplot of Termite Survival by Dose",
       xlab = "Number of Surviving Termites",
       ylab = "Dose (mg)",
       panel = function(x, y) {
           panel.bwplot(x, y)  # Default box plot
           panel.stripplot(x, y, jitter.data = TRUE, col = "red", pch = 16)  # Add points with jitter
       })



```

## 2.C)

```{r}

```

## 2.D)

```{r}

```


## 2.E)

```{r}

```

## 2.F)

```{r}

```