---
title:  
    "Week 6 exercise solutions"
date: 
    "Oct. 14th, 2024"
author:  
    "Ali Movasati, Isabelle Cretton, Tristan Koning"  
output:  
    pdf_document:
        latex_engine: xelatex
---

```{r global options}

# Set global code chunk options
knitr::opts_chunk$set(warning = FALSE)

```

```{r libraries}


# load required libraries
library(skimr)
library(ggplot2)
library(lme4)
```

# Exercise 1

## (a)

```{r}
# Load data
hearing <-  read.table(file = "day6/data/hearing.txt", sep = "\t", header = TRUE)
hearing <- within(hearing, {
  ListID <- factor(ListID, levels = c("List1", "List2", "List3", "List4"))
})
skim(hearing)

# Graphical summeries
pairs(hearing)

# Boxplot of scores by ListID
bxp1 <- ggplot(hearing, aes(x = ListID, y = Hearing, color = ListID)) +
  geom_boxplot() +
  labs(title = "Boxplot of Scores by ListID", x = "ListID", y = "Score")

# Scatterplot of subjectids on hearing scores
scat1 <- ggplot(hearing, aes(x = SubjectID, y = Hearing, color = ListID)) +
  geom_point() +
  labs(title = "Scatterplot of Scores by SubjectID", x = "SubjectID", y = "Score")

print(bxp1)
print(scat1)
```

## (b)
```{r}
model_A <- lm(Hearing ~ ListID, data = hearing)

summary(model_A)
anova(model_A)
```

We can explain 11% of the variance in hearing scores by the list ID variable (Adjusted R squared). 
Looking at the anova output, ListID has a significant p-value meaning that there is a significant difference in means between the lists.

## (c)
```{r}
# Fit a linear mixed model
model_B <- lmer(Hearing ~ (1 | SubjectID) + ListID, data = hearing)
summary(model_B)

plot(model_B)
confint(model_B, method = "boot", nsim = 100, oldNames = FALSE)
```

we cannot entirely conclude that the hearing scores differ for different lists, as there does exist an overlap in the confidence intervals of the estimates. 

## (d)
```{r}
# TODO
```