---
title: 'Exercise_1 solution'
date: 'Sept. 16th, 2024'
author: 'Tristan Koning'
output: pdf_document
---

```{r}
#install.packages("ggdendro")
library(ggplot2)
library(ggdendro)
library(cluster)
par(mfrow = c(1, 1))
```

# Problem 1

## (a)

```{r}
# Setup data for clustering
protein <- read.csv("day3/data/protein.txt", sep = "\t", header = TRUE)
row.names(protein) <- protein$Country
protein <- protein[, -1]
protein <- scale(protein)
```

```{r}
# Single Linkage Clustering
# Setup
single_linkage <- hclust(dist(protein), method = "single")

# Convert to dendrogram object
plot(single_linkage, main = "Single Linkage Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

```{r}
# Complete Linkage Clustering
# Setup
complete_linkage <- hclust(dist(protein), method = "complete")

plot(complete_linkage, main = "Complete Linkage Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

```{r}
# Ward Method Clustering
# Setup
ward_linkage <- hclust(dist(protein), method = "ward.D")

plot(ward_linkage, main = "Ward Method Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

- We can observe that Ward method clustering shows the clearest structure in the dendrogram. It groups countries with small distances together early, and then forms larger clusters incrementally. 

- The other two methods have a more erratic structure, with single linkage clustering showing the most erratic structure, where some countries dont belong to a cluster until the very end.

## (b)

```{r}
plot(ward_linkage, main = "Ward Method Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
rect.hclust(ward_linkage, k = 5, border = "red")
```

Looking at the dendrogram of the Ward method, we can observe that the countries are getting grouped together by their geographic location. We can split the dendrogram into 5 clusters, which are shown in the plot by the red borders. The clusters are as follows: Scandinavia, Eastern Europe, Western Europe, The Balkans, and the Mediterranean.

## (c)
```{r}
# Perform PCA on data
pca <- prcomp(protein, scale = TRUE)

# Get loadings of PC
loadings <- pca$rotation
loadings_data <- data.frame(Variable = rownames(loadings), PC1 = loadings[, 1], PC2 = loadings[, 2])

# Perform Dimensionality reduction
scores <- pca$x
pca_data <- data.frame(Country = data$Country, PC1 = scores[, 1], PC2 = scores[, 2])

ggplot() + 
  geom_point(data = pca_data, aes(x = PC1, y = PC2)) + 
  geom_text_repel(data = pca_data, aes(x = PC1, y = PC2, label = Country)) +
  geom_segment(data = loadings_data, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  geom_text_repel(data = loadings_data, aes(x = PC1, y = PC2, label = Variable), color = "red") +
  labs(title = "PCA Biplot of Protein Data", x = "PC1", y = "PC2") +
  theme_minimal()
```

Looking at the plot of the Dimensionality-Reduction and the dendogram, we can observe that they build similar clusters, which are based on the geographic regions of europe, Although the PCA plot has more of a tendency to show Western vs. Eastern Europe in a socio-economic context, while the dendogram shows more of a geographic clustering.
