---
title: 'Exercise 3 solution'
date: 'Oct. 1st, 2024'
author: 'Tristan, Köning, Ali Movasati, Isabelle Cretton'
output: pdf_document
---

```{r global options}
# Set global code chunk options
knitr::opts_chunk$set(warning = FALSE)
```

```{r libraries}
library(cluster)
library(stats)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tibble)
library(maps)
library(fields)
par(mfrow = c(1, 1))
```

# Problem 1 (Clustering)
## 1.A
```{r}
# Setup data for clustering
protein <- read.csv("day3/data/protein.txt", sep = "\t", header = TRUE)
row.names(protein) <- protein$Country
protein <- protein[, -1]
protein <- scale(protein)
```

```{r}
# Single Linkage Clustering
# Setup
single_linkage <- hclust(dist(protein), method = "single")

# Convert to dendrogram object
plot(single_linkage, main = "Single Linkage Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

```{r}
# Complete Linkage Clustering
# Setup
complete_linkage <- hclust(dist(protein), method = "complete")

plot(complete_linkage, main = "Complete Linkage Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

```{r}
# Ward Method Clustering
# Setup
ward_linkage <- hclust(dist(protein), method = "ward.D")

plot(ward_linkage, main = "Ward Method Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

- We can observe that Ward method clustering shows the clearest structure in the dendrogram. It groups countries with small distances together early, and then forms larger clusters incrementally. 

- The other two methods have a more erratic structure, with single linkage clustering showing the most erratic structure, where some countries dont belong to a cluster until the very end.

## 1.B
```{r}

# perform clustering
for (i in c("Country_clustering")){
    if (i == "Country_clustering"){
        data <- protein
    } else if (i == "Protein_clustering"){
        data <- protein_t
    }
    distance_matrix <- dist(data, method = "euclidean")
    hc_single <- hclust(distance_matrix, method = "ward.D")
    plot(hc_single, main = paste0("Dendrogram - Ward Linkage Method \n for ", i), xlab = "Data Points", ylab = "Height")
    abline(h = 70, col = "red", lwd = 2, lty = 2)
    abline(h = 38, col = "blue", lwd = 2, lty = 2)
}

```

```{r}
dev.off()
```

#### << Comments >>

We chose the clustering method Ward. If we cut the dendogram at distance of 70 we will end up with two clear-cut clusters. One of these clusters mostly contain the advanced European countries while the other one mostly contain the developing European countries.
Therefore, countries protein consumption patterns tend to reflect their underlying socio-economic.

