---
title: 'Exercise 3 solution'
date: 'Oct. 1st, 2024'
author: 'Tristan, Köning, Ali Movasati, Isabelle Cretton'
output: pdf_document
---

```{r global options}
# Set global code chunk options
knitr::opts_chunk$set(warning = FALSE)
```

```{r libraries}
library(cluster)
library(stats)
library(ggplot2)
library(ggrepel)
library(magrittr)
library(dplyr)
library(tibble)
library(maps)
library(fields)
par(mfrow = c(1, 1))
```

# Problem 1 (Clustering)
## 1.A
```{r}
# Setup data for clustering
protein <- read.csv("day3/data/protein.txt", sep = "\t", header = TRUE)
row.names(protein) <- protein$Country
protein <- protein[, -1]
protein <- scale(protein)
```

```{r}
# Single Linkage Clustering
# Setup
single_linkage <- hclust(dist(protein), method = "single")

# Convert to dendrogram object
plot(single_linkage, main = "Single Linkage Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

```{r}
# Complete Linkage Clustering
# Setup
complete_linkage <- hclust(dist(protein), method = "complete")

plot(complete_linkage, main = "Complete Linkage Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

```{r}
# Ward Method Clustering
# Setup
ward_linkage <- hclust(dist(protein), method = "ward.D")

plot(ward_linkage, main = "Ward Method Clustering Dendrogram", xlab = "Countries", ylab = "Euclidean Distance")
```

#### << Comments >>

- We can observe that Ward method clustering shows the clearest structure in the dendrogram. It groups countries with small distances together early, and then forms larger clusters incrementally. 

- The other two methods have a more erratic structure, with single linkage clustering showing the most erratic structure, where some countries dont belong to a cluster until the very end.

## 1.B
```{r}

# perform clustering
for (i in c("Country_clustering")){
    if (i == "Country_clustering"){
        data <- protein
    } else if (i == "Protein_clustering"){
        data <- protein_t
    }
    distance_matrix <- dist(data, method = "euclidean")
    hc_single <- hclust(distance_matrix, method = "ward.D")
    plot(hc_single, main = paste0("Dendrogram - Ward Linkage Method \n for ", i), xlab = "Data Points", ylab = "Height")
    abline(h = 70, col = "red", lwd = 2, lty = 2)
    abline(h = 38, col = "blue", lwd = 2, lty = 2)
}


rect.hclust(ward_linkage, k = 5, border = "red")
rect.hclust(ward_linkage, k = 2, border = "blue")
```

#### << Comments >>

- We can choose to look at either two or five groups, shown by the blue and red lines respectively. These can be representative of the economical status of the countries, with the two groups representing developed and developing countries, and the five groups being representative of different regions.

## 1.C
```{r}
# Perform PCA on data
protein_pca <- scale(protein)

# Perform PCA
pca <- prcomp(protein_pca, scale = TRUE)

# Get loadings of PC
loadings <- pca$rotation
loadings_data <- data.frame(Variable = rownames(loadings), PC1 = loadings[, 1], PC2 = loadings[, 2])

# Perform Dimensionality reduction
scores <- pca$x
pca_data <- data.frame(Country = rownames(protein), PC1 = scores[, 1], PC2 = scores[, 2])


ggplot() + 
  geom_point(data = pca_data, aes(x = PC1, y = PC2), color = "blue") + 
  geom_text_repel(data = pca_data, aes(x = PC1, y = PC2, label = Country)) +
  geom_segment(data = loadings_data, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = grid::arrow(length = unit(0.3, "cm")), color = "red") +
  geom_text_repel(data = loadings_data, aes(x = PC1, y = PC2, label = Variable), color = "red") +
  labs(title = "PCA Biplot of Protein Data", x = "PC1", y = "PC2") +
  theme_minimal()
```

#### << Comments >>

Looking at the plot of the Dimensionality-Reduction and the dendogram, we can observe that they build similar clusters, which are based on the geographic regions of europe, Although the PCA plot has more of a tendency to show Western vs. Eastern Europe in a socio-economic context, while the dendogram shows more of a geographic clustering.

# Problem 2 (my.kmeans)
```{r}
my.kmean <- function(x, k, iter = 10) {
    set.seed(111) 
    centroids <- sample(x, k, replace = FALSE)
    clusters <- numeric(length(x))
    
    for (i in 1:iter) {
        for (j in 1:length(x)) {
            distances <- abs(x[j] - centroids)
            clusters[j] <- which.min(distances)
        }
        for (c in 1:k) {
            centroids[c] <- mean(x[clusters == c])
        }
    }
    result <- data.frame(x = x, cluster = clusters)
    return(result)
}
x <- c(1,2,1,3,2,6,5,7,6,12)
k <- 3
result <- my.kmean(x, k)
result
```
