---
title:  
    "Day9 exercise solutions"
date: 
    "Nov. 15th, 2024"
author:  
    "Ali Movasati, Isabelle Caroline Rose Cretton, Tristan Koning"  
output:  
    pdf_document:
        latex_engine: xelatex
header-includes:
  - \usepackage{pdfpages}
---


```{r global options}

# Set global code chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

```


```{r libraries}


# load required libraries
library("skimr")
library("dplyr")
library("magrittr")
library("ggplot2")
library("survival")
library("survminer")
library("gridExtra")

# define functions
`%notin%` <- Negate(`%in%`)


```

# Problem 1

```{r}
log_logistic_density <- function(x, alpha, beta) {
    numerator <- (beta / alpha) * (x / alpha) ^ (beta - 1)
    denominator <- (1 + (x / alpha) ^ beta) ^ 2
    return(numerator / denominator)
}

log_logistic_cdf <- function(x, alpha, beta) {
    numerator <- x ^ beta
    denominator <- alpha ^ beta + x ^ beta	
    return(numerator / denominator)
}

log_logistic_survivor <- function(x, alpha, beta) {
    return(1 - log_logistic_cdf(x, alpha, beta))
}

log_logistic_hazard <- function(x, alpha, beta) {
    return(log_logistic_density(x, alpha, beta) / log_logistic_survivor(x, alpha, beta))
}
```

## (a)

```{r}
alpha <- 1
beta <- c(0.25, 0.5, 1, 4, 8)
x <- seq(0, 10, 0.1)

# Density function
density_data <- expand.grid(x = x, beta = beta) %>%
  mutate(density = log_logistic_density(x, alpha, beta))

p1 <- ggplot(density_data, aes(x = x, y = density, color = as.factor(beta))) +
  geom_line() +
  labs(title = "Log-logistic Density Function", x = "x", y = "Density", color = "Beta") +
  theme_minimal()

# CDF
cdf_data <- expand.grid(x = x, beta = beta) %>%
  mutate(density = log_logistic_cdf(x, alpha, beta))

p2 <- ggplot(cdf_data, aes(x = x, y = density, color = as.factor(beta))) +
    geom_line() +
    labs(title = "Log-logistic CDF", x = "x", y = "CDF", color = "Beta") +
    theme_minimal()

# Survivor function
survivor_data <- expand.grid(x = x, beta = beta) %>%
  mutate(density = log_logistic_survivor(x, alpha, beta))

p3 <- ggplot(survivor_data, aes(x = x, y = density, color = as.factor(beta))) +
    geom_line() +
    labs(title = "Log-logistic Survivor Function", x = "x", y = "Survivor Function", color = "Beta") +
    theme_minimal()

# Hazard function
hazard_data <- expand.grid(x = x, beta = beta) %>%
  mutate(density = log_logistic_hazard(x, alpha, beta))

p4 <- ggplot(hazard_data, aes(x = x, y = density, color = as.factor(beta))) +
    geom_line() +
    labs(title = "Log-logistic Hazard Function", x = "x", y = "Hazard Function", color = "Beta") +
    theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## (b)

```{r}
sigma <- 1
k <- c(0.5, 1, 2)
x <- seq(0, 10, 0.1)

# k1
density_data <- expand.grid(x = x, k = k) %>%
  mutate(density = dweibull(x, k, sigma))

ggplot(density_data, aes(x = x, y = density, color = as.factor(k))) +
    geom_line() + 
    labs(title = "Weibull Density Function", x = "x", y = "Density", color = "k") +
    theme_minimal()
```

# Problem 2

## (a)

```{r}
pbc_data <- read.table("data/pbc.txt", header = TRUE)
pbc_data$treat <- factor(pbc_data$treat, levels = c(1, 2), labels = c("Placebo", "Treatment"))
skim(pbc_data)

pbc_data$surv <- Surv(pbc_data$time, pbc_data$cenc0)
head(pbc_data$surv)
```

## (b)

```{r}
sfit <- survfit(surv ~ treat, data = pbc_data)
summary(sfit)

sfit_data <- as.data.frame(sfit$time)
sfit_data$surv <- sfit$surv
sfit_data$strata <- rep(names(sfit$strata), sfit$strata)

ggsurvplot(
  sfit,
  data = pbc_data,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  risk.table.col = "strata",
  linetype = "strata",
  palette = c("red", "blue"),
  xlab = "Time",
  ylab = "Survival Probability",
  title = "Survival Probability by Treatment",
  ggtheme = theme_minimal()
)

```

We cannot see a clear difference in survival between the two treatments. In fact, the Placebo group even has a higher Survival Probability for a while.

## (c)

```{r}
sdiff <- survdiff(surv ~ treat, data = pbc_data)
sdiff
```

We get a p-value of 1, which means that there is no significant difference in survival between the two treatments.

## (d)

```{r}
cox_model <- coxph(surv ~ treat, data = pbc_data)
summary(cox_model)
```

We get a p-value of 1, which means that there is no significant difference in survival between the two treatments.

## (e)

```{r}
cox_model2 <- coxph(surv ~ treat + age + cenc0, data = pbc_data)
summary(cox_model2)
```

By adding both age and cenc0 as predictors, we get a highly significant p-value in both the Likelihood ratio and logrank test, but surprisingly, the p-value for the Wald test is 0.9. This could be due to the fact that we have a small sample size.
We can also observe a high Concordance of 0.949, which indicates that the model is good at predicting the data.

The hazard rate for utilizing the treatment is by a factor of 0.944 smaller than without the treatment, for age it is by a factor of 1.014 and for cenc0 it is 1.181.

## (f)

```{r}
# Model 1: Only treatment
cox.zph(cox_model)
plot(cox.zph(cox_model))

# Model 2: Treatment, age and cenc0
par(mfrow = c(2, 2))
cox.zph(cox_model2)
plot(cox.zph(cox_model2))
```
