---
title:  
    "Day9 exercise solutions"
date: 
    "Nov. 13th, 2024"
author:  
    "Ali Movasati"  
output:  
    pdf_document:
        latex_engine: xelatex
header-includes:
  - \usepackage{pdfpages}
---




```{r global options}

# Set global code chunk options
knitr::opts_chunk$set(warning = FALSE)

```


```{r load libraries}


# load required libraries
library("skimr")
library("dplyr")
library("magrittr")
library("ggplot2")
library("survival")
library("survminer")
library("fields")

# define functions
`%notin%` <- Negate(`%in%`)


```

# Problem 1

## 1.A)

```{r}

load("/Users/alimos313/Documents/studies/phd/university/courses/stat-modelling/StatModelEx/day11/data/spatialSim.RData")

par(mfrow=c(1,2))
image.plot(sim1)
image.plot(sim2)

# summary statistics
summary(as.vector(sim1))
summary(as.vector(sim2))

```

## 1.B)

\includepdf[pages=-]{/Users/alimos313/Documents/studies/phd/university/courses/stat-modelling/StatModelEx/day11/sketch.pdf}

## 1.C)

```{r}


coords <- expand.grid(x1 = 1:100, x2 = 1:100)
coords$observed <- as.vector(sim1)

lm_model <- lm(observed ~ x1 + x2, data=coords)
summary(lm_model)


```



## 1.D)

```{r}
par(mfrow=c(2,2))
plot(lm_model)

```

### << comments >>

1. Assumption of Independent Errors
Violation: In spatial data, observations are often spatially correlated; nearby locations tend to have similar values 
(spatial autocorrelation). This violates the assumption that residuals (errors) are independent of each other.

2. Assumption of Homoscedasticity (Constant Variance of Errors)
Violation: Spatial data often exhibit heteroscedasticity, where the variability of the residuals changes across 
the spatial domain. This can occur if different areas have different levels of variability due to local conditions.

3. Assumption of Linearity
Violation: The relationship between predictors and the response variable might not be strictly linear across space. 
Spatial data often exhibit complex relationships that vary over the spatial domain, introducing non-linear patterns.

4. Assumption of Normality of Errors
Violation: The spatial structure or clusters in the data can lead to non-normal error distributions. 
Clustering or outliers common in spatial data might skew the residuals.

5. Collinearity Among Predictors
Violation: Spatial predictors (e.g., longitude, latitude, or environmental variables) are often correlated 
due to spatial patterns or regional influences. This can introduce multicollinearity.

## 1.E)

### << comments >>

A spatial regression model like kriging or spatial mixed models could be used:
Z(s) = μ + S(s) + ε
Where:

Z(s): Spatial process
μ: Mean trend
S(s): Spatial random effect
ε: Independent error term

# Problem 2

## 2.A)

```{r}

transect <- read.table(file = "/Users/alimos313/Documents/studies/phd/university/courses/stat-modelling/StatModelEx/day11/data/transect.txt", header = T, sep = " ", stringsAsFactor = F)


# Plot the data
plot(
  transect$s, transect$Z.s,
  type = "b",                      # Connect points with lines ("b" = both points and lines)
  pch = 16,                        # Use filled circles for points
  col = "blue",                    # Color of points and lines
  xlim = c(0, 10),                 # Set x-axis range to [0, 10]
  ylim = range(transect$Z.s),          # Automatically adjust y-axis range
  xlab = "Location (s)",           # Label for x-axis
  ylab = "Observation (Z(s))",     # Label for y-axis
  main = "Spatial Observations Along a Transect"
)

# Optionally add a grid for better readability
grid()
```


## 2.B)

```{r}
# Define the exponential covariance function
exp.cov <- function(h, theta2, theta3) {
  theta2 * exp(-h / theta3)
}

# Parameters
theta2 <- 5    # Sill
theta3 <- 2    # Range
h <- seq(0, 5, length.out = 100)  # Distances from 0 to 5

# Compute covariance values
cov_values <- exp.cov(h, theta2, theta3)

# Plot
plot(
  h, cov_values,
  type = "l",              # Line plot
  col = "blue",            # Line color
  lwd = 2,                 # Line width
  xlab = "Distance (h)",   # X-axis label
  ylab = "Covariance",     # Y-axis label
  main = "Exponential Covariance Function"
)

# Highlight key parameters
abline(h = theta2, col = "red", lty = 2)  # Sill (horizontal line)
abline(v = theta3, col = "green", lty = 2)  # Range (vertical line)
legend("topright", legend = c("Covariance", "Sill", "Range"),
       col = c("blue", "red", "green"), lty = c(1, 2, 2), bty = "n")

```


### << comments >>

Sill (θ2=5θ2​=5):
The red dashed line at 5 corresponds to the sill. It represents the maximum covariance value.
Range (θ3=2θ3​=2):
The green dashed line at h=2h=2 indicates the range, the distance at which the covariance drops to approximately θ2⋅e−1≈1.84θ2​⋅e−1≈1.84.
Nugget (θ1=0θ1​=0):
Since there’s no nugget effect, the covariance starts at the sill (cov(0)=θ2=5cov(0)=θ2​=5).


## 2.C)

```{r}

# Function to calculate pairwise Euclidean distances
dist.matrix <- function(x, y) {
  # Use outer() to calculate pairwise absolute differences
  outer(x, y, function(a, b) abs(a - b))
}

# Locations along the transect
locations <- c(1, 2, 4, 5, 7)

# Calculate the distance matrix
DIST.MAT <- dist.matrix(locations, locations)

# Print the distance matrix
print(DIST.MAT)
```


## 2.D)

```{r}


# Transect locations
locations <- c(1, 2, 4, 5, 7)

# Calculate the distance matrix
DIST.MAT <- dist.matrix(locations, locations)

# Parameters for the covariance function
theta2 <- 5    # Sill
theta3 <- 2    # Range

# Calculate the covariance matrix
SIGMA <- exp.cov(DIST.MAT, theta2, theta3)

# Print the covariance matrix
print(SIGMA)

```


## 2.E)

```{r}


# Observed locations
locations <- c(1, 2, 4, 5, 7)

# New locations where prediction is needed
snew <- seq(0, 10, length.out = 5) # For example, 5 evenly spaced points between 0 and 10

# Calculate distances between observed points and new locations
DIST.OBS.NEW <- dist.matrix(locations, snew)

# Parameters for the covariance function
theta2 <- 5    # Sill
theta3 <- 2    # Range

# Calculate the covariance matrix sigma
sigma <- exp.cov(DIST.OBS.NEW, theta2, theta3)

# Print results
cat("New locations (snew):\n")
print(snew)

cat("\nDistances between observed locations and snew (DIST.OBS.NEW):\n")
print(DIST.OBS.NEW)

cat("\nCovariance matrix (sigma):\n")
print(sigma)


```


## 2.F)

```{r}



# Observed locations and corresponding values
locations <- c(1, 2, 4, 5, 7)
Z_obs <- c(-2, 5, 1, -4, 2)  # Observed values of Z at locations s

# New locations where prediction is needed
snew <- seq(0, 10, length.out = 5) # Example: 5 evenly spaced points between 0 and 10

# Calculate pairwise distances
DIST.OBS.NEW <- dist.matrix(locations, snew)
DIST.OBS.OBS <- dist.matrix(locations, locations)

# Parameters for the covariance function
theta2 <- 5    # Sill
theta3 <- 2    # Range

# Calculate covariance matrices
C_OBS_NEW <- exp.cov(DIST.OBS.NEW, theta2, theta3)  # Covariance vector between observed and new
C_OBS_OBS <- exp.cov(DIST.OBS.OBS, theta2, theta3)  # Covariance matrix for observed locations

# Solve kriging system to calculate weights
C_OBS_OBS_inv <- solve(C_OBS_OBS)  # Inverse of the covariance matrix for observed points
weights <- C_OBS_NEW %*% C_OBS_OBS_inv  # Kriging weights

# Kriging prediction for the new locations
Z_pred <- weights %*% Z_obs  # Predicted values for the new locations

# Plot the observed and predicted values
plot(locations, Z_obs, col = "blue", pch = 16, cex = 1.5, xlab = "Location", ylab = "Z(s)", 
     main = "Kriging Prediction for Z(s)")

# Plot the predicted values
lines(snew, Z_pred, col = "red", lwd = 2)

# Optionally, plot points for predicted locations
points(snew, Z_pred, col = "red", pch = 17, cex = 1.5)


```


## 2.G)

```{r}


# Define parameters for the covariance function
theta2_values <- c(5, 10, 15)  # Different sill values to compare
theta3 <- 2                    # Range value

# Generate distances between 0 and 5
h <- seq(0, 5, length.out = 100)

# Plot the covariance function for different theta2 values
plot(h, exp.cov(h, theta2_values[1], theta3), type = "l", col = "red", lwd = 2,
     xlab = "Distance (h)", ylab = "Covariance", 
     main = "Exponential Covariance Function for Different θ2 Values")
lines(h, exp.cov(h, theta2_values[2], theta3), col = "blue", lwd = 2)
lines(h, exp.cov(h, theta2_values[3], theta3), col = "green", lwd = 2)
legend("topright", legend = paste("θ2 =", theta2_values), col = c("red", "blue", "green"), lwd = 2)

# Observed locations and values
locations <- c(1, 2, 4, 5, 7)
Z_obs <- c(-2, 5, 1, -4, 2)

# New prediction locations
snew <- seq(0, 10, length.out = 5)


# Create a plot to store the results
plot(locations, Z_obs, col = "blue", pch = 16, cex = 1.5, xlab = "Location", ylab = "Z(s)", 
     main = "Kriging Prediction for Z(s) with Different θ2 Values")

# Loop to calculate and plot results for different theta2 values
for (theta2 in theta2_values) {
  
  # Calculate covariance matrices for the given theta2
  C_OBS_NEW <- exp.cov(DIST.OBS.NEW, theta2 = theta2, theta3 = theta3)  # Covariance between observed and new
  C_OBS_OBS <- exp.cov(DIST.OBS.OBS, theta2 = theta2, theta3 = theta3)  # Covariance matrix for observed locations

  # Solve the kriging system
  C_OBS_OBS_inv <- solve(C_OBS_OBS)  # Inverse of the covariance matrix for observed points
  weights <- C_OBS_NEW %*% C_OBS_OBS_inv  # Kriging weights

  # Kriging prediction for the new locations
  Z_pred <- weights %*% Z_obs  # Predicted values for the new locations

  # Plot the predicted values as a line
  lines(snew, Z_pred, col = ifelse(theta2 == 5, "red", ifelse(theta2 == 10, "green", "purple")), lwd = 2)
  
  # Optionally, plot points for predicted locations
  points(snew, Z_pred, col = ifelse(theta2 == 5, "red", ifelse(theta2 == 10, "green", "purple")), pch = 17, cex = 1.5)
}

# Add a legend to the plot
legend("topright", legend = paste("θ2 =", theta2_values), col = c("red", "green", "purple"), lwd = 2)

```


Effect on the covariance function: As θ2θ2​ increases, the covariance between locations increases, 
implying a stronger spatial correlation.

Effect on kriging predictions: We do not see a difference on kriging predictions. 
It's likely because the range of the predicted values for different θ2θ2​ values isn't significantly different. 
This can happen if the locations where the predictions are made (the snew locations) are not very far apart, or if the model's sensitivity to θ2θ2​ is low for the given parameters.