---
title: "Week 7 Solutions"
author: "Isabelle Cretton"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
#install.packages("sm")
#install.packages("splines")
library(sm)  # for kernel regression
library(splines)  # for smoothing splines
```

## Problem 1: Non-parametric regression

### Data Loading and Initial Setup

```{r load-data}
# Read the data
medflies <- read.table("day7/data/medflies.txt", header = TRUE)
head(medflies)
```

### (a) Data Exploration

```{r exploration}
# Summary statistics
summary(medflies)

# Create plots to visualize the data
p1 <- ggplot(medflies, aes(x = day, y = living)) +
  geom_line() +
  labs(title = "Number of Living Flies Over Time",
       x = "Day",
       y = "Number of Living Flies")

p2 <- ggplot(medflies, aes(x = day, y = mort.rate)) +
  geom_line() +
  labs(title = "Mortality Rate Over Time",
       x = "Day",
       y = "Mortality Rate")

print(p1)
print(p2)
```

The data shows the survival and mortality patterns of Mediterranean fruit flies over time. The number of living flies decreases over time, while the mortality rate shows more complex patterns.

### (b) Mortality Rate Verification

```{r verify-mort-rate}
# Calculate mortality rate from the 'living' variable
medflies$calc_mort_rate <- c(-diff(medflies$living) / medflies$living[-nrow(medflies)], NA)

# Compare calculated vs. provided mortality rates
head(data.frame(
  day = medflies$day,
  original_rate = medflies$mort.rate,
  calculated_rate = medflies$calc_mort_rate
))

# Plot comparison
ggplot(medflies, aes(x = mort.rate, y = calc_mort_rate)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(title = "Comparison of Original vs. Calculated Mortality Rates",
       x = "Original Mortality Rate",
       y = "Calculated Mortality Rate")
```

### (c) Verification of Gompertz's Theory

```{r gompertz}
# Ensure mort.rate is numeric and handle log transformation properly
medflies$mort.rate <- as.numeric(medflies$mort.rate)
# Add small constant to avoid log(0)
medflies$log_mort_rate <- log(medflies$mort.rate + 1e-10)

# Fit linear regression on log-transformed data
model_full <- lm(log_mort_rate ~ day, data = medflies)

# Plot
ggplot(medflies, aes(x = day, y = log_mort_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Log Mortality Rate vs. Day",
       x = "Day",
       y = "Log Mortality Rate")

# Diagnostic plots
par(mfrow = c(2,2))
plot(model_full)

# Try finding a subset where theory holds better
# Let's look at days 20-40
subset_data <- subset(medflies, day >= 20 & day <= 40)
model_subset <- lm(log_mort_rate ~ day, data = subset_data)

ggplot(subset_data, aes(x = day, y = log_mort_rate)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Log Mortality Rate vs. Day (Days 20-40)",
       x = "Day",
       y = "Log Mortality Rate")
```

### (d) Kernel Regression

```{r kernel}
# Try different kernels and bandwidths
h_values <- c(2, 5, 10)
kernels <- c("normal", "epanech", "triangular")

# Function to plot kernel regression with different parameters
plot_kernel <- function(h, kernel) {
  sm.regression(medflies$day, 
               medflies$mort.rate,
               h = h,
               kernel = kernel,
               display = "none") -> fit
  
  data.frame(x = fit$eval.points,
             y = fit$estimate) -> smooth_data
  
  ggplot() +
    geom_point(data = medflies, aes(x = day, y = mort.rate), alpha = 0.3) +
    geom_line(data = smooth_data, aes(x = x, y = y), color = "red") +
    labs(title = paste("Kernel:", kernel, "Bandwidth:", h),
         x = "Day",
         y = "Mortality Rate")
}

# Plot different combinations
for(h in h_values) {
  for(k in kernels) {
    print(plot_kernel(h, k))
  }
}
```

### (e) Smoothing Splines and Local Polynomials

```{r splines-local}
# (i) Smoothing splines
smooth_spline <- smooth.spline(medflies$day, medflies$mort.rate)

# (ii) Local polynomials (using loess)
local_poly <- loess(mort.rate ~ day, data = medflies, span = 0.75)

# Plot both methods
ggplot(medflies, aes(x = day, y = mort.rate)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = predict(smooth_spline), color = "Smoothing Spline")) +
  geom_line(aes(y = predict(local_poly), color = "Local Polynomial")) +
  scale_color_manual(values = c("red", "blue")) +
  labs(title = "Comparison of Smoothing Methods",
       x = "Day",
       y = "Mortality Rate",
       color = "Method")
```

### (f) Cross-validation for Optimal Bandwidth

```{r cv-bandwidth}
# Find optimal bandwidth using cross-validation
h.cv <- hcv(medflies$day, medflies$mort.rate)

# Fit model with optimal bandwidth
sm.regression(medflies$day, 
             medflies$mort.rate,
             h = h.cv,
             display = "none") -> optimal_fit

# Plot result
ggplot() +
  geom_point(data = medflies, aes(x = day, y = mort.rate), alpha = 0.3) +
  geom_line(data = data.frame(x = optimal_fit$eval.points,
                             y = optimal_fit$estimate),
            aes(x = x, y = y), color = "red") +
  labs(title = paste("Kernel Regression with CV-optimal bandwidth:", round(h.cv, 2)),
       x = "Day",
       y = "Mortality Rate")
```

### (g) Manual Cross-validation for Smoothing Splines

```{r manual-cv}
# Function to compute CV score for smoothing splines
cv_spline <- function(df) {
  cv_scores <- numeric(length(medflies$mort.rate))
  
  for(i in 1:length(medflies$mort.rate)) {
    # Fit model without i-th observation
    fit <- smooth.spline(medflies$day[-i], 
                        medflies$mort.rate[-i], 
                        df = df)
    # Predict i-th observation
    pred <- predict(fit, medflies$day[i])$y
    # Compute squared error
    cv_scores[i] <- (medflies$mort.rate[i] - pred)^2
  }
  
  mean(cv_scores)
}

# Try different degrees of freedom
df_values <- seq(4, 20, by = 2)
cv_results <- sapply(df_values, cv_spline)

# Plot CV scores
plot(df_values, cv_results, type = "b",
     xlab = "Degrees of Freedom",
     ylab = "CV Score",
     main = "Cross-validation Scores for Different df")

# Optimal df
opt_df <- df_values[which.min(cv_results)]
```

### (h) Scientific Questions

Non-parametric models are useful for:
- Understanding the general pattern of mortality rates over time without assuming a specific functional form
- Identifying periods of unusual mortality rate changes
- Describing the day-to-day variation in mortality rates

Linear models (like in Gompertz's theory) are useful for:
- Testing specific theories about exponential growth in mortality rates
- Making predictions about mortality rates at specific ages
- Quantifying the rate of increase in mortality over time

The choice between parametric and non-parametric models depends on the scientific question at hand and the assumptions we're willing to make about the underlying relationship between variables.