---
title: "Exercise 5 Solutions"
author: "Isabelle Cretton"
output: 
  pdf_document:
    latex_engine: xelatex
--- 

```{r global options}
# Set global code chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r setup, include=FALSE}
# Load necessary libraries
#install.packages("car")
library(car)
#install.packages("lmtest")

# Load the package
library(lmtest)
```

## Problem 1: Multiple linear regression – teacher salaries

### 1.A Read and Process Data
```{r}
# Read the data
salary_data <- read.table("day5/data/salary.txt", header = TRUE, sep = ",", quote = "\"")

# Create a factor variable for district size
salary_data$size <- factor(salary_data$districtSize, 
                           levels = 1:3, 
                           labels = c("< 1000 students", "1000 - 2000 students", "> 2000 students"))

head(salary_data)
```


### 1.B Numerical and Graphical Summaries
```{r}
# Numerical summaries
summary(salary_data)

# Average salary by district size
aggregate(salary ~ size, data = salary_data, mean)

# Graphical summaries
par(mfrow = c(2, 2))
plot(salary ~ experience, data = salary_data, main = "Salary vs Experience")
boxplot(salary ~ size, data = salary_data, main = "Salary by District Size")
hist(salary_data$salary, main = "Histogram of Salaries", xlab = "Salary")
pairs(salary_data[, c("salary", "experience", "districtSize")], main = "Scatterplot Matrix")

# Calculate average salary in CHF (assuming 1 USD = 0.89 CHF as of October 2023)
mean_salary_chf <- mean(salary_data$salary) * 0.89
print(paste("Average salary in CHF:", round(mean_salary_chf, 2)))
```

- There's a positive correlation between salary and experience.
- Salary tends to increase with district size.
- There's no clear relationship between experience and district size.


### 1.C Fit and Compare Models
```{r}
# Fit models
model_A <- lm(salary ~ experience + districtSize, data = salary_data)
model_B <- lm(salary ~ experience + size, data = salary_data)

# Compare models
summary(model_A)
summary(model_B)

# Compare model fit
anova(model_A, model_B)
```

### Comment 
*Model A*: salary ~ experience + districtSize
*Model B*: salary ~ experience + size (as a factor)
Both models show similar performance:

- *Model A*: Adjusted R-squared = 0.5748
- *Model B*: Adjusted R-squared = 0.5741

The ANOVA test comparing the two models shows no significant difference (p-value = 0.5145), indicating that using district size as a factor (Model B) doesn't significantly improve the model fit compared to using it as a continuous variable (Model A).

### 1.D Model B Discussion
Model B is statistically significant (F-statistic: 146.6, p-value < 2.2e-16).
R-squared of 0.5741 suggests moderate explanatory power.
All variables are highly significant (p-values < 2e-16):

Experience: $584.15 increase per year
Medium districts: $3,088 higher than small districts
Large districts: $5,732.28 higher than small districts

### 1.E Modified Model B
```{r}
# Fit the modified Model B
model_B_modified <- lm(salary ~ I(experience - 13) + size, data = salary_data)

# Compare summaries
summary(model_B)
summary(model_B_modified)

# Compare coefficients
cbind(coef(model_B), coef(model_B_modified))
```

The modified Model B shows:
1. The intercept changed from 24995.49 to 32589.46.
2. he coefficients for experience and district size remained the same:
    - Experience: 584.15 (unchanged)
    - Medium-sized districts: 3088.00 (unchanged)
    - Large districts: 5732.28 (unchanged)

### 1.F Check Regression Assumptions
```{r}
# Residual plots
par(mfrow = c(2, 2))
plot(model_B)

# Additional diagnostic tests
library(car)

# Normality of residuals
shapiro.test(residuals(model_B))
qqPlot(model_B, main="QQ Plot")

# Homoscedasticity
ncvTest(model_B)
spreadLevelPlot(model_B)

# Multicollinearity
vif(model_B)

# Influential observations
influencePlot(model_B, id.method="identify", main="Influence Plot", sub="Circle size is proportional to Cook's Distance")

# Durbin-Watson test for autocorrelation
dwtest(model_B)
```
- Linearity: Reasonably met (random scatter in Residuals vs Fitted plot)
- Homoscedasticity: Slight heteroscedasticity observed
- Normality: Approximately normal with some tail deviations
- Influential observations: A few potential influential points (e.g., 511, 317, 173)
Assumptions are reasonably met with some potential issues.

### 1.G Salary Prediction
```{r}
# Create a new data point
new_teacher <- data.frame(experience = 10, 
                          districtSize = 3, 
                          size = "> 2000 students")

# Predict using Model A
predict(model_A, newdata = new_teacher, interval = "confidence")

# Predict using Model B
predict(model_B, newdata = new_teacher, interval = "confidence")
```
For a teacher with 10 years of experience in a large district:

Model A prediction:
- Point estimate: $36,704.42
- 95% Confidence Interval: ($36,040.00, $37,368.84)

Model B prediction:
- Point estimate: $36,569.29
- 95% Confidence Interval: ($35,789.40, $37,349.17)

Both models give similar predictions, with Model A predicting a slightly higher salary. The confidence intervals overlap substantially, indicating that the predictions are not significantly different between the two models.