---
title: "Week 8 Solutions"
author: "Isabelle Caroline Rose Cretton"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(faraway)  # for the bliss dataset
```

## Problem 1: Logistic Regression

### (a) Data Exploration

```{r data-exploration}
# Load the data
data(bliss, package = "faraway")

# Calculate total insects and ratio
bliss$ratio <- bliss$dead / (bliss$dead + bliss$alive)

# Plot the ratio vs concentration
ggplot(bliss, aes(x = conc, y = ratio)) +
  geom_point() +
  labs(title = "Ratio of Dead Insects vs. Insecticide Concentration",
       x = "Concentration",
       y = "Proportion of Dead Insects") +
  theme_minimal()
```

### (b) Logistic Regression with Logit Link

```{r logit-regression}
# Fit logistic regression
logit_model <- glm(cbind(dead, alive) ~ conc, 
                   family = binomial(link = "logit"), 
                   data = bliss)

# Display results
summary(logit_model)
```

### (c) Manual Prediction Calculation

```{r manual-prediction}
# Calculate predicted values manually
coeffs <- coef(logit_model)
linear_pred <- coeffs[1] + coeffs[2] * bliss$conc
manual_pred <- exp(linear_pred) / (1 + exp(linear_pred))

# Compare with fitted values
fitted_pred <- fitted(logit_model)

# Compare results
comparison_df <- data.frame(
  concentration = bliss$conc,
  manual = manual_pred,
  fitted = fitted_pred,
  difference = abs(manual_pred - fitted_pred)
)
print(comparison_df)
```

### (d) Add Predicted Probabilities to Plot

```{r predicted-plot}
# Create prediction grid
pred_grid <- data.frame(conc = seq(min(bliss$conc), max(bliss$conc), length.out = 100))
pred_grid$pred <- predict(logit_model, newdata = pred_grid, type = "response")

ggplot() +
  geom_point(data = bliss, aes(x = conc, y = ratio)) +
  geom_line(data = pred_grid, aes(x = conc, y = pred), color = "blue") +
  labs(title = "Observed Proportions and Predicted Probabilities (Logit)",
       x = "Concentration",
       y = "Proportion of Dead Insects") +
  theme_minimal()
```

### (e) Add Confidence Intervals

```{r confidence-intervals}
# Calculate confidence intervals
pred_ci <- predict(logit_model, 
                  newdata = pred_grid, 
                  type = "link", 
                  se.fit = TRUE)

# Transform to probability scale
ci_lower <- plogis(pred_ci$fit - 1.645 * pred_ci$se.fit)  # 90% CI
ci_upper <- plogis(pred_ci$fit + 1.645 * pred_ci$se.fit)

# Add to plot
pred_grid$lower <- ci_lower
pred_grid$upper <- ci_upper

ggplot() +
  geom_point(data = bliss, aes(x = conc, y = ratio)) +
  geom_line(data = pred_grid, aes(x = conc, y = pred), color = "blue") +
  geom_ribbon(data = pred_grid, 
              aes(x = conc, ymin = lower, ymax = upper),
              alpha = 0.2) +
  labs(title = "Logistic Regression with 90% Confidence Interval",
       x = "Concentration",
       y = "Proportion of Dead Insects") +
  theme_minimal()
```

### (f) Probit Link Function

```{r probit-regression}
# Fit probit model
probit_model <- glm(cbind(dead, alive) ~ conc, 
                    family = binomial(link = "probit"), 
                    data = bliss)

# Get predictions
pred_grid$probit_pred <- predict(probit_model, 
                                newdata = pred_grid, 
                                type = "response")

# Calculate confidence intervals
probit_ci <- predict(probit_model, 
                    newdata = pred_grid, 
                    type = "link", 
                    se.fit = TRUE)

pred_grid$probit_lower <- pnorm(probit_ci$fit - 1.645 * probit_ci$se.fit)
pred_grid$probit_upper <- pnorm(probit_ci$fit + 1.645 * probit_ci$se.fit)

# Plot both models
ggplot() +
  geom_point(data = bliss, aes(x = conc, y = ratio)) +
  geom_line(data = pred_grid, aes(x = conc, y = pred, color = "Logit")) +
  geom_line(data = pred_grid, aes(x = conc, y = probit_pred, color = "Probit")) +
  geom_ribbon(data = pred_grid, 
              aes(x = conc, ymin = lower, ymax = upper),
              alpha = 0.1, fill = "blue") +
  geom_ribbon(data = pred_grid, 
              aes(x = conc, ymin = probit_lower, ymax = probit_upper),
              alpha = 0.1, fill = "red") +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "Comparison of Logit and Probit Models",
       x = "Concentration",
       y = "Proportion of Dead Insects",
       color = "Model") +
  theme_minimal()
```

### (g) Compare Logit and Probit Predictions

```{r model-comparison}
# Create comparison grid
comp_grid <- data.frame(conc = seq(-1, 5, length.out = 100))

# Get predictions
comp_grid$logit_pred <- predict(logit_model, 
                               newdata = comp_grid, 
                               type = "response")
comp_grid$probit_pred <- predict(probit_model, 
                                newdata = comp_grid, 
                                type = "response")
comp_grid$ratio <- comp_grid$logit_pred / comp_grid$probit_pred

ggplot(comp_grid, aes(x = conc, y = ratio)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(title = "Ratio of Logit to Probit Predictions",
       x = "Concentration",
       y = "Logit/Probit Ratio") +
  theme_minimal()
```

### (h) Calculate LD50

```{r ld50}
# For logit model, LD50 is when logit = 0
coeffs <- coef(logit_model)
ld50 <- -coeffs[1] / coeffs[2]
print(paste("LD50 (concentration at which 50% die):", round(ld50, 3)))
```

## Problem 2: Exponential Family

The exponential family has the form:

$f(y; \theta, \phi) = exp((y\theta - b(\theta))/\phi + c(y, \phi))$

### (a) Exponential Distribution

The probability density function is:
$f(y; \lambda) = \lambda e^{-\lambda y}$

We can rewrite this as:
$f(y; \lambda) = exp(log(\lambda) - \lambda y) = exp(-\lambda y + log(\lambda))$

This belongs to the exponential family with:

* $\theta = -\lambda$
* $\phi = 1$
* $b(\theta) = -log(-\theta)$
* $c(y, \phi) = 0$

### (b) Binomial Distribution

The probability density function is:
$f(y; \pi) = {n \choose y}\pi^y(1-\pi)^{n-y}$

We can rewrite this as:
$f(y; \pi) = exp(y log(\pi/(1-\pi)) + n log(1-\pi) + log({n \choose y}))$
$\phantom{f(y; \pi)} = exp(y\theta - n log(1 + e^\theta) + log({n \choose y}))$

where $\theta = log(\pi/(1-\pi))$

This belongs to the exponential family with:

* $\theta = log(\pi/(1-\pi))$
* $\phi = 1$
* $b(\theta) = n log(1 + e^\theta)$
* $c(y, \phi) = log({n \choose y})$

### (c) Uniform Distribution

The probability density function is:
$f(y; \theta) = 1/\theta, \quad 0 < y < \theta$

This cannot be written in exponential family form because the support of $y$ depends on $\theta$.

### (d) Normal Distribution (known variance)

The probability density function is:
$f(y; \mu, \sigma^2) = (1/\sqrt{2\pi\sigma^2})exp(-(y-\mu)^2/(2\sigma^2))$

We can rewrite this as:
$f(y; \mu, \sigma^2) = exp(y\mu/\sigma^2 - \mu^2/(2\sigma^2) - y^2/(2\sigma^2) - (1/2)log(2\pi\sigma^2))$

This belongs to the exponential family with:

* $\theta = \mu$
* $\phi = \sigma^2$
* $b(\theta) = \theta^2/2$
* $c(y, \phi) = -y^2/(2\phi) - (1/2)log(2\pi\phi)$