---
title:  
    "Day8 exercise solutions"
date: 
    "Oct. 4th, 2024"
author:  
    "Ali Movasati, Isabelle Caroline Rose Cretton, Tristan Koning"  
output:  
    pdf_document:
        latex_engine: xelatex
header-includes:
  - \usepackage{pdfpages}
---


```{r global options}

# Set global code chunk options
knitr::opts_chunk$set(warning = FALSE)

```


```{r libraries}


# load required libraries
library("skimr")
library("dplyr")
library("magrittr")
library("ggplot2")

# define functions
`%notin%` <- Negate(`%in%`)


```

# Problem 1

```{r load the  data}

data(bliss, package = "faraway")
bliss %<>% mutate(ratio = dead/(alive + dead))

```

## 1.A)

```{r table summary}

skim(bliss)
head(bliss)

```


```{r graphic summary}
# Load the data
data(bliss, package = "faraway")

# Calculate total insects and ratio
bliss$ratio <- bliss$dead / (bliss$dead + bliss$alive)

# Plot the ratio vs concentration
ggplot(bliss, aes(x = conc, y = ratio)) +
  geom_point() +
  labs(title = "Ratio of Dead Insects vs. Insecticide Concentration",
       x = "Concentration",
       y = "Proportion of Dead Insects") +
  theme_minimal()
```


## 1.B)

```{r}

logit_model <- glm(cbind(dead,alive)~conc, family = binomial(link=logit), data = bliss)

summary(logit_model)

```



## 1.C)

```{r manual-prediction}
# Calculate predicted values manually
coeffs <- coef(logit_model)
linear_pred <- coeffs[1] + coeffs[2] * bliss$conc
manual_pred <- exp(linear_pred) / (1 + exp(linear_pred))

# Compare with fitted values
fitted_pred <- fitted(logit_model)

# Compare results
comparison_df <- data.frame(
  concentration = bliss$conc,
  manual = manual_pred,
  fitted = fitted_pred,
  difference = abs(manual_pred - fitted_pred)
)
print(comparison_df)
```



## 1.D)

```{r predicted-plot}
# Create prediction grid
pred_grid <- data.frame(conc = seq(min(bliss$conc), max(bliss$conc), length.out = 100))
pred_grid$pred <- predict(logit_model, newdata = pred_grid, type = "response")

ggplot() +
  geom_point(data = bliss, aes(x = conc, y = ratio)) +
  geom_line(data = pred_grid, aes(x = conc, y = pred), color = "blue") +
  labs(title = "Observed Proportions and Predicted Probabilities (Logit)",
       x = "Concentration",
       y = "Proportion of Dead Insects") +
  theme_minimal()
```

## 1.E)

```{r confidence-intervals}
# Calculate confidence intervals
pred_ci <- predict(logit_model, 
                  newdata = pred_grid, 
                  type = "link", 
                  se.fit = TRUE)

# Transform to probability scale
ci_lower <- plogis(pred_ci$fit - 1.645 * pred_ci$se.fit)  # 90% CI
ci_upper <- plogis(pred_ci$fit + 1.645 * pred_ci$se.fit)

# Add to plot
pred_grid$lower <- ci_lower
pred_grid$upper <- ci_upper

ggplot() +
  geom_point(data = bliss, aes(x = conc, y = ratio)) +
  geom_line(data = pred_grid, aes(x = conc, y = pred), color = "blue") +
  geom_ribbon(data = pred_grid, 
              aes(x = conc, ymin = lower, ymax = upper),
              alpha = 0.2,
              fill = "blue") +
  labs(title = "Logistic Regression with 90% Confidence Interval",
       x = "Concentration",
       y = "Proportion of Dead Insects") +
  theme_minimal()
```


## 1.F)

```{r probit-regression}
# Fit probit model
probit_model <- glm(cbind(dead, alive) ~ conc, 
                    family = binomial(link = "probit"), 
                    data = bliss)

# Get predictions
pred_grid$probit_pred <- predict(probit_model, 
                                newdata = pred_grid, 
                                type = "response")

# Calculate confidence intervals
probit_ci <- predict(probit_model, 
                    newdata = pred_grid, 
                    type = "link", 
                    se.fit = TRUE)

pred_grid$probit_lower <- pnorm(probit_ci$fit - 1.645 * probit_ci$se.fit)
pred_grid$probit_upper <- pnorm(probit_ci$fit + 1.645 * probit_ci$se.fit)


```


## 1.G)

```{r visualize both models}

# Plot both models
ggplot() +
  geom_point(data = bliss, aes(x = conc, y = ratio)) +
  geom_line(data = pred_grid, aes(x = conc, y = pred, color = "Logit")) +
  geom_line(data = pred_grid, aes(x = conc, y = probit_pred, color = "Probit")) +
  geom_ribbon(data = pred_grid, 
              aes(x = conc, ymin = lower, ymax = upper),
              alpha = 0.1, fill = "blue",
              fill = "blue") +
  geom_ribbon(data = pred_grid, 
              aes(x = conc, ymin = probit_lower, ymax = probit_upper),
              alpha = 0.1, fill = "red") +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "Comparison of Logit and Probit Models",
       x = "Concentration",
       y = "Proportion of Dead Insects",
       color = "Model") +
  theme_minimal()
```



## 1.H)


```{r}


logit_model_out <- predict(logit_model, newdata=data.frame(conc=seq(-1,to=5,length.out=100000)), type="response")
ld50 <- round(seq(-1,to=5,length.out=100000)[which.min(abs(logit_model_out-0.5))])

print(paste0("LD50 or the dose at which there is 50% chance of survival is ", ld50))

```


## Problem 2: Exponential Family

The exponential family has the form:

$f(y; \theta, \phi) = exp((y\theta - b(\theta))/\phi + c(y, \phi))$

### (a) Exponential Distribution

The probability density function is:
$f(y; \lambda) = \lambda e^{-\lambda y}$

We can rewrite this as:
$f(y; \lambda) = exp(log(\lambda) - \lambda y) = exp(-\lambda y + log(\lambda))$

This belongs to the exponential family with:

* $\theta = -\lambda$
* $\phi = 1$
* $b(\theta) = -log(-\theta)$
* $c(y, \phi) = 0$

### (b) Binomial Distribution

The probability density function is:
$f(y; \pi) = {n \choose y}\pi^y(1-\pi)^{n-y}$

We can rewrite this as:
$f(y; \pi) = exp(y log(\pi/(1-\pi)) + n log(1-\pi) + log({n \choose y}))$
$\phantom{f(y; \pi)} = exp(y\theta - n log(1 + e^\theta) + log({n \choose y}))$

where $\theta = log(\pi/(1-\pi))$

This belongs to the exponential family with:

* $\theta = log(\pi/(1-\pi))$
* $\phi = 1$
* $b(\theta) = n log(1 + e^\theta)$
* $c(y, \phi) = log({n \choose y})$

### (c) Uniform Distribution



The probability density function is:
$f(y; \theta) = 1/\theta, \quad 0 < y < \theta$

This cannot be written in exponential family form because the support of $y$ depends on $\theta$.



### (d) Normal Distribution (known variance)



The probability density function is:
$f(y; \mu, \sigma^2) = (1/\sqrt{2\pi\sigma^2})exp(-(y-\mu)^2/(2\sigma^2))$

We can rewrite this as:
$f(y; \mu, \sigma^2) = exp(y\mu/\sigma^2 - \mu^2/(2\sigma^2) - y^2/(2\sigma^2) - (1/2)log(2\pi\sigma^2))$



* **Natural parameter**: \( \theta = \frac{\mu}{\sigma^2} \)
* **Dispersion parameter**: \( \phi = \sigma^2 \)
* **Function \( b(\theta) \)**: \( b(\theta) = \frac{\mu^2}{2 \sigma^2} \)
* **Function \( c(y, \phi) \)**: \( c(y, \phi) = -\frac{y^2}{2 \phi^2} - \frac{1}{2} \ln(2 \pi \phi^2) \)