---
title: 'Exercise_1 solution'
date: 'Sept. 16th, 2024'
author: 'Tristan Koning'
output: pdf_document
---

```{r}
install.packages("ggrepel")
install.packages("ggplot2")
install.packages("ggbiplot")
library(ggplot2)
library(ggrepel)
library(ggbiplot)
```

# Problem 1

## (a)

```{r}
# Setup
data <-  read.csv("day2/data/protein.txt", sep = "\t", header = TRUE)

# Perform PCA on data, except country names
pca <- prcomp(data[, -1], scale = TRUE)

# Variance explained by each PC
screeplot(pca)
round(cumsum(pca$sdev^2 / sum(pca$sdev^2)), 3)

# Get loadings of PC
loadings <- pca$rotation
loadings_data <- data.frame(Variable = rownames(loadings), PC1 = loadings[, 1], PC2 = loadings[, 2])

# Perform Dimensionality reduction
scores <- pca$x
pca_data <- data.frame(Country = data$Country, PC1 = scores[, 1], PC2 = scores[, 2])

ggplot() + 
  geom_point(data = pca_data, aes(x = PC1, y = PC2)) + 
  geom_text_repel(data = pca_data, aes(x = PC1, y = PC2, label = Country)) +
  geom_segment(data = loadings_data, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  geom_text_repel(data = loadings_data, aes(x = PC1, y = PC2, label = Variable), color = "red") +
  labs(title = "PCA Biplot of Protein Data", x = "PC1", y = "PC2") +
  theme_minimal()
```

Looking at the Biplot of the PCA together with its Loadings, we can observe that Eastern European Countries consume their protein more along PC1, which primarily includes Nuts and Cereals.
On the other Hand, Western European and Scandinavian Countries consume their protein via Meat, Eggs and Milk.
Mediterranean Latin Countries such as Portugal, Spain and Greece consume theirs along both PC1 and 2, which includs Fruites, Vegetables and Fish additionally to what has already been mentioned along PC1.

## (b)
Given the high cumulative variance explained by the first 4 components, and diminishing returns from including more components, it would be reasonable to focus on the first 4 components for interpreting and understanding the data.
Breakdown of what each component represents:
- PC1 (44.52%): likely captures the general protein consumption pattern, with higher loadings on food groups such as Fish and Fr&Veg (Fruits & Vegetables). Countries like Portugal and Spain have high scores on PC1 due to higher fish consumption, while countries like Romania consume more cereals and have lower PC1 scores.

- PC2 (18.17%): may reflect meat consumption patterns, particularly distinguishing countries that consume more Red Meat (like the UK and France) from those with lower consumption of these food groups.

- PC3 (12.53%): could represent a trade-off between White Meat and Dairy (Milk) consumption, differentiating countries like Ireland and W Germany from those with lower dairy consumption.

- PC4 (10.61%): might capture more specific dietary differences related to nuts, eggs, and starch, distinguishing countries like Greece and Hungary from others.

# Problem 2
## (a)

```{r}
# Setup
require(maps)
require(fields)
load("day2/data/prec_jan_feb.RData")

```

```{r}
# Plot maps of Jamuary 1st
image.plot(lon, lat, pre[, , 1, main = "January 1st, 00:00"])
map("world", add = TRUE)
image.plot(lon, lat, pre[, , 2], main = "January 1st, 06:00")
map("world", add = TRUE)
image.plot(lon, lat, pre[, , 3,  main = "January 1st, 12:00"])
map("world", add = TRUE)
image.plot(lon, lat, pre[, , 4], main = "January 1st, 18:00")
map("world", add = TRUE)
```

```{r}
# Plot mean and variance
image.plot(lon, lat, apply(pre, c(1, 2), mean), main = "Mean")
map("world", add = TRUE)

image.plot(lon, lat, apply(pre, c(1, 2), var), main = "Variance")
map("world", add = TRUE)
```

## (b)

```{r}
# Data matrix: 31 x 19 spatial locations, 240 observations
pre2 <- t(array(pre, c(dim(pre)[1] * dim(pre)[2], dim(pre)[3])))

# Perform PCA
pca <- prcomp(pre2, scale = TRUE)

# Variance explained by each PC
screeplot(pca)
round(cumsum(pca$sdev^2 / sum(pca$sdev^2)), 3)

# Get loadings of top 6 PCs that explain the most variance
loadings <- pca$rotation[, 1:6]

# Plot for each loading
par(mfrow = c(2, 3))
for (i in 1:6) {
  image.plot(lon, lat, matrix(loadings[, i], nrow = 31), main = paste("PC", i))
  map("world", add = TRUE)
}
```

## (c)

```{r}
# Calculate Eigenvalues from sdev
eigenvalues <- pca$sdev^2

# Display Eigenvalues
par(mfrow = c(1, 1))
barplot(eigenvalues, main = "Eigenvalues", xlab = "PC", ylab = "Eigenvalue")
```

## (d)

```{r}
# Setup
n <- 240

for (i in 1:(length(eigenvalues) - 1)) {
  if ((eigenvalues[i] - eigenvalues[i+1]) / eigenvalues[i] < sqrt(2 / n)) {
    num_components <- i
    break
  }
}

# Show results
num_components
screeplot(pca[1:num_components])
round(cumsum(pca$sdev^2 / sum(pca$sdev^2)), 3)[1:num_components]
```

We truncated at 7 components, which results in 50.1% of the variance explained.

# Problem 3

## (a)

```{r}
# Setup
mu <- c(2, 5)
sigma <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)

# Perform Eigendecomposition
eig <- eigen(sigma)
```

Eigenvalues: `r eig$values`
Eigenvectors: `r eig$vectors`

## (b)

```{r}
# Setup
n <- 100
set.seed(123)

# Draw from multivariate normal
sample <- rmvnorm(n = n, mean = mu, sigma = sigma)

# Plot samples
data <- data.frame(x = sample[, 1], y = sample[, 2])
ggplot(data = data, aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Samples from Multivariate Normal Distribution")
```

## (c)

```{r}
# Setup eigenvectors to ensure they start at the mean of the distribution
eig_data <- data.frame(
  x = c(2, 2),
  y = c(5, 5),
  xend = 2 + eig$vectors[, 1],
  yend = 5 + eig$vectors[, 2]
)

# Plot with Eigenvectors
ggplot(data = data, aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Samples from Multivariate Normal Distribution") +
  geom_segment(data = eig_data, aes(x = x, y = y, xend = xend, yend = yend), arrow = arrow(length = unit(0.3, "cm")), color = "red")
```

## (d)

```{r}
# Perform PCA on sample
pca <- prcomp(sample, scale = TRUE)

eigenvalues_sample <- pca$sdev^2

# Display rotated points
rotated_points <- pca$x
rotated_data <- data.frame(x = rotated_points[, 1], y = rotated_points[, 2])
ggplot(data = rotated_data, aes(x = x, y = y)) +
  geom_point() +
  labs(title = "Rotated Points from PCA")
```

Original Eigenvalues: `r eig$values`
Sample Eigenvalues: `r eigenvalues_sample`

They are not exactly the same since we are dealing with a random sample.
This means that we will have some small variations in the data that will influence the eigenvalues. 
We can still observe that the sample eigenvalues are close to the original eigenvalues.