---
title: "02PCA ex STA121"
author: "Isabelle Caroline Rose Cretton"
date: "September 27, 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Data loading:
Load from https://user.math.uzh.ch/furrer/download/sta121/ the following data sets:
- elephant.csv
- Meatprice.txt
```{r}
# Load the data sets
# 1. elephant.csv
elephant <- read.table(file = "day2/data/elephant.csv", header = TRUE, sep = ",")
# 2. Meatprice.txt
meatprice <- read.table("day2/data/Meatprice.txt", header = TRUE, sep = "\t")
```

```{r}
View(elephant)
#View(meatprice)
```

```{r}
View(meatprice)
```

Have a look at the data, e.g. View(dataset) or head(dataset), plotting -> to see if reading the data worked properly.

Rdata   
https://user.math.uzh.ch/furrer/download/sta121/
Penguins.RData

```{r}
# "./data/Penguins.RData"

```


# 2. Working with data frames

Dataset Seatbelts is available in R from the package datasets. 
You should have access to this data set without loading.
Use ?Seatbelts to read about the dataset. 
First we will change the data set to be a data frame (not a Time series), see the code. 
```{r}
library(datasets)
Seatbelts_df <- data.frame(as.matrix(Seatbelts), date=time(Seatbelts)) 
View(Seatbelts_df)

# convert Time Series to data frame so that we can work with that

```
1. Add a column representing the ratio of rear passengers killed to front passengers killed
```{r}
```


