---
title: "02PCA ex STA121"
author: "Isabelle Caroline Rose Cretton"
date: "September 27, 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
#install.packages("ggbiplot")
library(ggbiplot)
#install.packages("fields")
library(fields)  # for image.plot()
#install.packages("maps")
library(maps)    # for adding world borders
#install.packages("reshape2")
library(reshape2)
```

# Problem 1: PCA on the protein data
# part a
```{r}
# Load the protein dataset
protein <- read.csv("day2/protein.txt", sep = "\t", header = TRUE)

# Set row names and remove the country column
rownames(protein) <- protein$Country
protein <- protein[,-1]

# Perform PCA
pca <- prcomp(protein, scale. = TRUE)

# Summary of PCA to check variance explained
pca_summary <- summary(pca)
print(pca_summary)
```
PC1 is the most important component, capturing 44.52% of the variance. This suggests that a large portion of the variability in protein consumption across the countries can be explained by a single underlying factor (likely related to major dietary differences, such as fish and cereal consumption patterns).
PC2 captures an additional 18.17%, indicating it adds another dimension to understanding the variation in the data.
PC3 and PC4 contribute further, and collectively, the first four components explain about 85.82% of the total variance in the data. This implies that using these four components, we can represent the majority of the structure in the data while keeping complexity manageable.

## Scree plot
```{r}
# Scree plot for the variance explained by each component
screeplot(pca, type = "lines", main = "Scree Plot")
```
The elbow rule suggests that after PC2 or PC3, the variance explained by each successive component decreases significantly.

## Biplot
```{r}
ggbiplot(pca, labels = rownames(protein), ellipse = TRUE, circle = TRUE) + 
  ggtitle("PCA Biplot for Protein Data") + theme_minimal()
```
This biplot indicates that protein consumption patterns are geographically distinct across European countries, with certain countries having unique dietary habits (e.g., high fish consumption in Portugal and Spain, and high cereal consumption in Romania and Bulgaria). The plot also shows that Cereals, Fish, and Fr&Veg are the most influential food categories in shaping these differences.
The x-axis (PC1) explains 44.5% of the total variance in the dataset, in turn, the y-axis (PC2) explains 25% of the variance.

```{r}
# Pairwise plot for the first four principal components
pairs(pca$x[,1:4], labels = c("PC1", "PC2", "PC3", "PC4"), 
      main = "Pairwise Plot of First Four Principal Components")
```
- PC1 vs. PC2: Shows some separation between countries, indicating these components capture the most significant differences in protein consumption patterns (e.g., high fish/fruit consumption vs. high cereal consumption).

- PC1 vs. PC3, PC1 vs. PC4: Some spread but no strong clustering, suggesting that PC3 and PC4 capture more subtle or nuanced variations.

- PC2 vs. PC3, PC2 vs. PC4: Similar to PC1, some spread but no clear group separation.

- PC3 vs. PC4: Little visible structure or separation, indicating these components explain smaller or independent variations.

## part b
Given the high cumulative variance explained by the first 4 components, and diminishing returns from including more components, it would be reasonable to focus on the first 4 components for interpreting and understanding the data.
Breakdown of what each component represents:
- PC1 (44.52%): likely captures the general protein consumption pattern, with higher loadings on food groups such as Fish and Fr&Veg (Fruits & Vegetables). Countries like Portugal and Spain have high scores on PC1 due to higher fish consumption, while countries like Romania consume more cereals and have lower PC1 scores.

- PC2 (18.17%): may reflect meat consumption patterns, particularly distinguishing countries that consume more Red Meat (like the UK and France) from those with lower consumption of these food groups.

- PC3 (12.53%): could represent a trade-off between White Meat and Dairy (Milk) consumption, differentiating countries like Ireland and W Germany from those with lower dairy consumption.

- PC4 (10.61%): might capture more specific dietary differences related to nuts, eggs, and starch, distinguishing countries like Greece and Hungary from others.

# Problem 2: PCA on the precipitable water data
```{r}
# Load the RData file
load("day2/prec_jan_feb.RData")

# Check dimensions of the 'pre' array, should be c(31, 19, 240)
dim(pre)

# Reshape the data for the first four time slices (January 1st, 00:00 to 18:00)
pre_melted <- melt(pre[, , 1:4])  # Melt the first 4 time points (00:00, 06:00, 12:00, 18:00)
colnames(pre_melted) <- c("LongitudeIndex", "LatitudeIndex", "Time", "PrecipitableWater")

# Create data frames for longitude and latitude
lon_df <- data.frame(LongitudeIndex = 1:length(lon), Longitude = lon)
lat_df <- data.frame(LatitudeIndex = 1:length(lat), Latitude = lat)

# Merge longitude and latitude with the melted data
pre_melted <- merge(pre_melted, lon_df, by = "LongitudeIndex")
pre_melted <- merge(pre_melted, lat_df, by = "LatitudeIndex")

# Update time labels for the four times on January 1st
pre_melted$Time <- factor(pre_melted$Time, labels = c("00:00", "06:00", "12:00", "18:00"))

# Plot with ggplot2
ggplot(pre_melted, aes(x = Longitude, y = Latitude, fill = PrecipitableWater)) +
  geom_raster() +
  borders("world", colour = "black") +  # Add country borders
  coord_fixed(xlim = c(-30, 50), ylim = c(30, 75)) +  # Focus on Europe (adjust limits)
  scale_fill_viridis_c(option = "C", limits = c(min(pre_melted$PrecipitableWater), 
                                                max(pre_melted$PrecipitableWater))) +  
  facet_wrap(~ Time) +  # Facet by time
  labs(title = "Precipitable Water on January 1st, 2000", fill = "Precipitable Water (mm)") +
  theme_minimal() +  # Minimal theme for clarity
  theme(
    axis.title = element_blank(),       # Remove axis titles
    strip.text = element_text(size = 12), # Larger facet titles
    plot.title = element_text(hjust = 0.5, size = 16),  # Center and increase the size of the main title
    legend.position = "right"           # Adjust legend position
  )
```

```{r}
# prepare to plot the mean and variance fields
mean_field <- apply(pre, c(1, 2), mean)
var_field <- apply(pre, c(1, 2), var)

# reshape the mean and variance fields for ggplot2
mean_melted <- melt(mean_field)
var_melted <- melt(var_field)
colnames(mean_melted) <- colnames(var_melted) <- c("LongitudeIndex", "LatitudeIndex", "Value")

#create data frames for longitude and latitude
lon_df <- data.frame(LongitudeIndex = 1:length(lon), Longitude = lon)
lat_df <- data.frame(LatitudeIndex = 1:length(lat), Latitude = lat)

# merge longitude and latitude with the melted mean and variance fields
mean_melted <- merge(mean_melted, lon_df, by = "LongitudeIndex")
mean_melted <- merge(mean_melted, lat_df, by = "LatitudeIndex")

var_melted <- merge(var_melted, lon_df, by = "LongitudeIndex")
var_melted <- merge(var_melted, lat_df, by = "LatitudeIndex")
```

```{r}
# Plot the mean field
mean_plot <- ggplot(mean_melted, aes(x = Longitude, y = Latitude, fill = Value)) +
  geom_raster() +
  borders("world", colour = "black") +  # Add country borders
  coord_fixed(xlim = c(-30, 50), ylim = c(30, 75)) +  # Focus on Europe
  scale_fill_viridis_c(option = "C", name = "Mean") +  # Color scale for mean
  labs(title = "Mean Field of Precipitable Water Data") +
  theme_minimal() +
  theme(axis.title = element_blank(), plot.title = element_text(hjust = 0.5, size = 16))
mean_plot
```

```{r}
# Plot the variance field
var_plot <- ggplot(var_melted, aes(x = Longitude, y = Latitude, fill = Value)) +
  geom_raster() +
  borders("world", colour = "black") +  # Add country borders
  coord_fixed(xlim = c(-30, 50), ylim = c(30, 75)) +  # Focus on Europe
  scale_fill_viridis_c(option = "C", name = "Variance") +  # Color scale for variance
  labs(title = "Variance Field of Precipitable Water Data") +
  theme_minimal() +
  theme(axis.title = element_blank(), plot.title = element_text(hjust = 0.5, size = 16))
var_plot
```

### part b
```{r}
# Flatten spatial dimensions (lon and lat) into one dimension
pre_reshaped <- array_reshape(pre, c(31 * 19, 240))  # 31 (lon) * 19 (lat) by 240 (time)

# Step 2: Perform PCA
pca_pre <- prcomp(pre_reshaped, scale. = TRUE)  # PCA on the reshaped data

# Step 3: Check how much variance is explained by each component
pca_summary <- summary(pca_pre)
print(pca_summary)
```

## plotting
```{r}
# The loadings for PC1 are stored in pca_pre$rotation[, 1] (First component's loadings)
# Reshape PC1 loadings into spatial format (31 longitude x 19 latitude)
pc1_map <- matrix(pca_pre$rotation[, 1], nrow = 31, ncol = 19)

#Reshape the PC1 map for plotting with ggplot2
pc1_melted <- melt(pc1_map)
colnames(pc1_melted) <- c("LongitudeIndex", "LatitudeIndex", "PC1_Loading")

# Create data frames for longitude and latitude
lon_df <- data.frame(LongitudeIndex = 1:length(lon), Longitude = lon)
lat_df <- data.frame(LatitudeIndex = 1:length(lat), Latitude = lat)

#Merge longitude and latitude with the melted PC1 loadings
pc1_melted <- merge(pc1_melted, lon_df, by = "LongitudeIndex")
pc1_melted <- merge(pc1_melted, lat_df, by = "LatitudeIndex")

#Check structure before plotting
str(pc1_melted)  # Ensure that columns Longitude, Latitude, and PC1_Loading exist

#Plot the first principal component map
ggplot(pc1_melted, aes(x = Longitude, y = Latitude, fill = PC1_Loading)) +
  geom_raster() +
  geom_contour(aes(z = PC1_Loading), color = "white", alpha = 0.5) +  # Add contour lines for better distinction
  borders("world", colour = "black") +  # Add country borders
  coord_fixed(xlim = c(-30, 50), ylim = c(30, 75)) +  # Focus on Europe
  scale_fill_viridis_c(option = "C", name = "PC1 Loading", limits = c(min(pc1_melted$PC1_Loading), max(pc1_melted$PC1_Loading))) +  # Tweak color range
  labs(title = "First Principal Component (PC1) Loadings with Contours") +
  theme_minimal() +
  theme(axis.title = element_blank(), plot.title = element_text(hjust = 0.5, size = 16))
```
PC1 explains 58.37% of the total variability in the data. This is a very high proportion, indicating that the first principal component captures most of the variance in the data. The corresponding map of PC1 loadings reflects the most significant pattern of variability in precipitable water content across the region.

## plotting PC2 and PC3
```{r}
# Step 1: Extract the second and third principal component (PC2, PC3) loadings
pc2_map <- matrix(pca_pre$rotation[, 2], nrow = 31, ncol = 19)  # Reshape PC2 loadings into spatial format
pc3_map <- matrix(pca_pre$rotation[, 3], nrow = 31, ncol = 19)  # Reshape PC3 loadings into spatial format

# Step 2: Reshape the PC2 and PC3 maps for ggplot2
pc2_melted <- melt(pc2_map)
colnames(pc2_melted) <- c("LongitudeIndex", "LatitudeIndex", "PC2_Loading")

pc3_melted <- melt(pc3_map)
colnames(pc3_melted) <- c("LongitudeIndex", "LatitudeIndex", "PC3_Loading")

# Step 3: Merge longitude and latitude with the melted PC2 and PC3 loadings
pc2_melted <- merge(pc2_melted, lon_df, by = "LongitudeIndex")
pc2_melted <- merge(pc2_melted, lat_df, by = "LatitudeIndex")

pc3_melted <- merge(pc3_melted, lon_df, by = "LongitudeIndex")
pc3_melted <- merge(pc3_melted, lat_df, by = "LatitudeIndex")
```

### plotting PC2
```{r}
ggplot(pc2_melted, aes(x = Longitude, y = Latitude, fill = PC2_Loading)) +
  geom_raster() +
  geom_contour(aes(z = PC2_Loading), color = "white", alpha = 0.5) +  # Add contour lines for better distinction
  borders("world", colour = "black") +  # Add country borders
  coord_fixed(xlim = c(-30, 50), ylim = c(30, 75)) +  # Focus on Europe
  scale_fill_viridis_c(option = "C", name = "PC2 Loading") +  # Color scale for PC2 loadings
  labs(title = "Second Principal Component (PC2) Loadings") +
  theme_minimal() +
  theme(axis.title = element_blank(), plot.title = element_text(hjust = 0.5, size = 16))
```

### plotting PC3
```{r}
ggplot(pc3_melted, aes(x = Longitude, y = Latitude, fill = PC3_Loading)) +
  geom_raster() +
  geom_contour(aes(z = PC3_Loading), color = "white", alpha = 0.5) +  # Add contour lines for better distinction
  borders("world", colour = "black") +  # Add country borders
  coord_fixed(xlim = c(-30, 50), ylim = c(30, 75)) +  # Focus on Europe
  scale_fill_viridis_c(option = "C", name = "PC3 Loading") +  # Color scale for PC3 loadings
  labs(title = "Third Principal Component (PC3) Loadings") +
  theme_minimal() +
  theme(axis.title = element_blank(), plot.title = element_text(hjust = 0.5, size = 16))
```
PC2 and PC3 provide more nuanced spatial patterns than PC1, each revealing specific types of variability that are not captured by the dominant PC1. While PC1 mostly captured the overall variance across Europe, PC2 and PC3 are picking up more regional and localized patterns.
The horizontal striations in PC2 suggest the importance of large-scale atmospheric patterns, whereas PC3’s localized features may reflect the influence of more regional weather systems or topography.

## Plotting PC1 & PC2, and PC1, PC2 & PC3
```{r}
# Combine the first three principal components' scores into a data frame
pc_scores <- as.data.frame(pca_pre$x[, 1:3])  # Extract PC1, PC2, PC3 scores
colnames(pc_scores) <- c("PC1", "PC2", "PC3")  # Rename columns

# Pairwise plot for PC1 and PC2
pairs(pc_scores[, 1:2],  # PC1 and PC2 only
      labels = c("PC1", "PC2"), 
      main = "Pairwise Plot of PC1 and PC2",
      col = "blue", pch = 16)

# Pairwise plot for PC1, PC2, and PC3
pairs(pc_scores,  # PC1, PC2, PC3 together
      labels = c("PC1", "PC2", "PC3"), 
      main = "Pairwise Plot of PC1, PC2, and PC3",
      col = "blue", pch = 16)
```
The pairwise plot of PC1, PC2, and PC3 reveals distinct patterns of variability captured by each principal component. While PC1 shows the most concentrated clustering, the relationships between PC2 and PC3 suggest more dispersed and orthogonal variability, indicating that each component adds unique dimensions to the data's structure.

## part c
```{r}
eigenvalues <- pca$sdev^2

par(mfrow = c(1, 1))  # Ensure single plot layout

# Create a bar plot and store the bar positions
bar_positions <- barplot(
  eigenvalues, 
  main = "Eigenvalues of Principal Components", 
  xlab = "Principal Component", 
  ylab = "Eigenvalue",
  col = "lightblue",  # Use a distinct color for bars
  border = "black",   # Black border around bars
  ylim = c(0, max(eigenvalues) * 1.1)  # Add some space at the top for labels
)

#Add eigenvalue labels on top of each bar
text(
  bar_positions, 
  eigenvalues, 
  labels = round(eigenvalues, 2),  # Round to 2 decimal places
  pos = 3,  # Position the text above the bars
  cex = 0.8,  # Text size
  col = "black"  # Text color
)
#Add axis labels with integer principal component indices
axis(1, at = bar_positions, labels = 1:length(eigenvalues), las = 2)  # X-axis labels
```

```{r}
plot(cumulative_variance, type = "b", pch = 19, col = "darkgreen",
     xlab = "Principal Component", ylab = "Cumulative Proportion of Variance Explained",
     main = "Cumulative Variance Explained",
     ylim = c(0, 1))
abline(h = 0.9, col = "red", lty = 2)  # Reference line for 90% variance explained
```
Based on both plots, approximately 15 to 20 principal components are relevant, as they explain around 90% of the variance. After that point, the additional components contribute minimally to the overall variance, making them less important for the analysis.

## part d
```{r}
n <- 240  # Total number of eigenvalues/observations

num_components <- length(eigenvalues)  # Initialize with max components
for (i in 1:(length(eigenvalues) - 1)) {
  if ((eigenvalues[i] - eigenvalues[i+1]) / eigenvalues[i] < sqrt(2 / n)) {
    num_components <- i
    break
  }
}
num_components
```

```{r}
variance_explained <- eigenvalues / sum(eigenvalues)
cumulative_variance_truncated <- round(cumsum(variance_explained)[1:num_components], 3)

# Create a bar plot for the scree plot
bar_positions <- barplot(
  variance_explained[1:num_components], 
  main = paste("Scree Plot (Truncated at", num_components, "PCs)"), 
  xlab = "Principal Component", 
  ylab = "Variance Explained",
  col = "lightblue",  
  border = "black",    
  ylim = c(0, max(variance_explained) * 1.1)
)

# Add labels on top of each bar
text(bar_positions, variance_explained[1:num_components], 
     labels = round(variance_explained[1:num_components], 2), 
     pos = 3, cex = 0.8, col = "black")

# Print cumulative variance for the retained components
print(cumulative_variance_truncated)
```
9 components are relevant based on the North et al. rule.


# Problem 3: PCA 
```{r}
# Load necessary library
library(MASS)

# Define the mean vector and covariance matrix
mu <- c(2, 5)
Sigma <- matrix(c(1, 1/2, 1/2, 1), nrow = 2)

# Calculate eigenvalues and eigenvectors
eigen_decomp <- eigen(Sigma)
eigenvalues <- eigen_decomp$values
eigenvectors <- eigen_decomp$vectors

# Output the eigenvalues and eigenvectors
print("Eigenvalues:")
print(eigenvalues)
print("Eigenvectors:")
print(eigenvectors)
```

### part b
```{r}
set.seed(123)  # Set seed for reproducibility
n <- 100
data <- mvrnorm(n, mu, Sigma)

# Create a scatter plot
plot(data, main = "Scatter Plot of 100 Realizations", xlab = "X1", ylab = "X2", pch = 19, col = "blue")
```

### part c
```{r}
arrows(2, 5, 2 + eigenvectors[1, 1] * eigenvalues[1], 5 + eigenvectors[2, 1] * eigenvalues[1], 
       col = "red", length = 0.1)
arrows(2, 5, 2 + eigenvectors[1, 2] * eigenvalues[2], 5 + eigenvectors[2, 2] * eigenvalues[2], 
       col = "green", length = 0.1)

# Add a legend
legend("topright", legend = c("Eigenvector 1", "Eigenvector 2"), col = c("red", "green"), lty = 1, bty = "n")
```

### part d
```{r}
pc_data <- scale(data) %*% eigenvectors  # Centering the data and projecting onto eigenvectors

# Plot the rotated points
plot(pc_data, main = "Principal Components", xlab = "PC1", ylab = "PC2", pch = 19, col = "purple")
```
#### Why eigenvalues in PCA not the same?
Eigenvalues from PCA are scaled by the variance of the data.
The eigenvalues of the covariance matrix represent variance in the direction of the eigenvectors.
